% acknowl.tex {Acknowledgements}

\chapter*{Acknowledgements}


We wish to express our deepest sense of gratitude and sincere thanks to all those who have contributed to the successful completion of this project report titled, \textbf{System Threat Forecaster} which was undertaken as a requirement for the AICTE QIP PG Certification Programme on "Deep Learning: Fundamentals and Applications."

We are profoundly indebted to the faculty members, especially our guide and mentor for this project, at the Department of Electronics Engineering at Sardar Vallabhbhai National Institute of Technology (SVNIT), Surat, Gujarat, for providing the excellent learning environment, technical facilities, and insightful direction that were crucial for this work.

We extend our sincere thanks to the leadership and organizing committee for the successful execution of the QIP course: \textbf{Prof. Anupam Shukla, Director, SVNIT Surat (Chair Patron)}, \textbf{Dr. Shilpi Gupta, HOD, DOECE, SVNIT Surat (Patron)}, \textbf{Prof. A. A. Shaikh, DoME, SVNIT, Surat (Centre Coordinator)}, \textbf{Dr. Jignesh N. Sarvaiya, Dr. Kishor Upla, Dr. Kamal Captain, and Dr. Suman Deb (Coordinators from DoECE)}.

We also acknowledge the support and sponsorship provided by the All India Council for Technical Education (AICTE) under the Quality Improvement Programme (QIP), which has enabled us to enhance our knowledge and skills in deep learning through this opportunity.

Finally, we express our gratitude to our respective parent institutions, \textbf{Government Polytechnic Palanpur}, for granting the necessary support and time to participate in this Program.

\vspace{1cm}

\begin{flushright} 
Mr. Milav Jayeshkumar Dabgar\\[0.01in]
Lecturer, Dept of ECE\\
GP Palanpur\\
DECEMBER 2025
\end{flushright}